---
title: "BOO 2024 - BMD analysis"
output: html_document
date: "`r Sys.Date()`"
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, ggpubr, corrr, ggfortify, ggcorrplot, ggdendro, data.table, GGally)
```



## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# input directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

# plot directory
if(!dir.exists("PLOT")){
  dir.create(path = file.path(getwd(), "PLOT"))
}
plot_dir <- file.path(getwd(), "PLOT")
```



# Load data {.tabset}

## Metadata
Here you should load the metadata you obtained.
* What is metadata and how many samples do you have?
The data with the characteristics of the samples, 12368
```{r}
metadata_fsample <- fread(input = file.path(input_dir, "20240521_metadata_processed.csv"))
```

## Countdata
Here you should load the raw data you obtained.
* What type of normalization was applied?
CPM normalization
```{r}
countdata_cpm <- fread(input = file.path(input_dir, "20240521_countdata_cpm_processed.csv"))
```


## Wrangle countdata for BMD analysis
```{r}
metadata_fsample <- metadata_fsample |>
  mutate(COMPOUND_TIME = paste0(COMPOUND,"_",TIME))

genes = countdata_cpm |> pull(GENE_SYMBOL)

compound = "CYCLOSPORINE A"

compound_time <- metadata_fsample |> 
  filter(COMPOUND == compound) |>
  pull(COMPOUND_TIME) |> 
  unique()
  

for(i in compound_time) {
  metadata_subset <- metadata_fsample |>
    filter(COMPOUND_TIME == i)
  
  countdata_subset = countdata_cpm |>
    select(GENE_SYMBOL,metadata_subset$SAMPLE_ID) |>
    column_to_rownames("GENE_SYMBOL") |>
    t() |>
    as.data.frame() |>
    rownames_to_column("SAMPLE_ID") |>
    left_join(metadata_subset |> dplyr::select(SAMPLE_ID,COMPOUND,CONCENTRATION,DOSE_LEVEL,REPLICATE)) |>
    arrange(DOSE_LEVEL,REPLICATE) |>
    dplyr::select(SAMPLE_ID, CONCENTRATION, all_of(genes)) |>
    column_to_rownames("SAMPLE_ID") |>
    t() |>
    as.data.frame() |>
    rownames_to_column("GENE_SYMBOL")
  
  fwrite(countdata_subset, paste0(input_dir, "/BMD_input_",i,"countdata_normalized_gene.txt"), row.names = F, sep = "\t")
}
```



## Load data in BMDExpress 2.3
Here you will load and analyse the data in BMDExpress 2.3 using the import function of a Generic geneset with a BASE2 transformation. 

First perform a William's Trend test, with p<0.05 threshold, but without any fold change threshold.
* How many genes do show a dose / concentration response for each of the timeponts?
4: 201
8: 750
16: 4641
24: 4943
48: 710
72: 1123

Then perform a BMD analysis with the standard settings. 
You can save the output in a text file. 
* Which five genes have the lowest BMD for each timepoint.
4: RBSN, BIRC2, NAIP, AGAP6, DDX39B
8: XIAP, INPP5B, NEGR1, EPHB2, LATS2
16: PWWP3B, NIPAL1, TAF5L, TMEM35B, SCTR
24: TENT5B, AHDC1, KBTBD12, TAGLN3, SERPINA4
48: ARMCX4, BMERB1, GALNT2, OSR2, LTO1
72: ORAI1, COX8A, TUBA1B, UGT1A8, AMMECR1L

Finally perform a pathway analysis, first with REACTOME and secondly with a predefined geneset from the RPTEC/TERT1 modules.
* Which pathways show the most significant concentration responses?
* Which modules show the most significant concentration responses?
* Are these two results pointing at similar pathways?

```{r}
library(readr)
library(dplyr)
library(ggplot2)

input_dir <- "C:/Users/Lenovo/Documents/BOO_CDS_2024"

# Load the datasets
Modules_data <- read_delim(file.path(input_dir, "Category_Analyses_filtered.txt"), skip = 4, delim = "\t")
Wiki_data <- read_delim(file.path(input_dir, "Wiki_Analyses_filtered.txt"), skip = 4, delim = "\t")
KEGG_data <- read_delim(file.path(input_dir, "KEGG_Analyses_filtered.txt"), skip = 4, delim = "\t")


```

```{r}
bmd_sd_1 <- Modules_data$`BMD Standard Deviation`
bmd_sd_2 <- Wiki_data$`BMD Standard Deviation`
bmd_sd_3 <- KEGG_data$`BMD Standard Deviation`

bmd_sd <- list(data1 = bmd_sd_1, data2 = bmd_sd_2, data3 = bmd_sd_3)

boxplot(bmd_sd, 
        main = "Comparison of BMD Standard Deviation",
        names = c("Modules", "Wiki", "KEGG"),
        ylab = "BMD Standard Deviation")
```

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Extract time point from Analysis column
data_kegg <- KEGG_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.factor(gsub(".*_(\\d+)countdata.*", "\\1", Analysis)))

# Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways <- data_kegg %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

# Ensure only the selected pathways are included
pathway_list <- unique(top_pathways$`GO/Pathway/Gene Set/Gene Name`)
filtered_data <- data_kegg %>% filter(`GO/Pathway/Gene Set/Gene Name` %in% pathway_list)

# Add a new column to represent the range for the error bars
filtered_data <- filtered_data %>%
  mutate(lower = `BMD Median` - `BMD Standard Deviation`,
         upper = `BMD Median` + `BMD Standard Deviation`)

# Plot the BMD Median trends
p_1 <- ggplot(filtered_data, aes(x = timepoint, y = `BMD Median`, color = `GO/Pathway/Gene Set/Gene Name`)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "BMD Median Trends of KEGG Pathways Over Time",
       x = "Timepoint",
       y = "BMD Median",
       color = "KEGG Pathway") +
  theme_minimal() +
  facet_wrap(~`GO/Pathway/Gene Set/Gene Name`) +
  theme(legend.position = "none",
        strip.text = element_text(size = 5)) 

# Save the plot with desired dimensions
ggsave("BMD_Median_Trends_KEGG.png", plot = p_1, width = 12, height = 8)
```

```{r}
# Extract time point from Analysis column
data_wiki <- Wiki_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.factor(gsub(".*_(\\d+)countdata.*", "\\1", Analysis)))

# Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways <- data_wiki %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

# Ensure only the selected pathways are included
pathway_list <- unique(top_pathways$`GO/Pathway/Gene Set/Gene Name`)
filtered_data <- data_wiki %>% filter(`GO/Pathway/Gene Set/Gene Name` %in% pathway_list)

# Add a new column to represent the range for the error bars
filtered_data <- filtered_data %>%
  mutate(lower = `BMD Median` - `BMD Standard Deviation`,
         upper = `BMD Median` + `BMD Standard Deviation`)

# Plot the BMD Median trends
p_2 <- ggplot(filtered_data, aes(x = timepoint, y = `BMD Median`, color = `GO/Pathway/Gene Set/Gene Name`)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "BMD Median Trends of Wiki Pathways Over Time",
       x = "Timepoint",
       y = "BMD Median",
       color = "Wiki Pathway") +
  theme_minimal() +
  facet_wrap(~`GO/Pathway/Gene Set/Gene Name`) +
  theme(legend.position = "none",
        strip.text = element_text(size = 6))  

# Save the plot with desired dimensions
ggsave("BMD_Median_Trends_Wiki.png", plot = p_2, width = 12, height = 8)
```

```{r}
# Extract time point from Analysis column
data_modules <- Modules_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.factor(gsub(".*_(\\d+)countdata.*", "\\1", Analysis)))

# Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways <- data_modules %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

# Ensure only the selected pathways are included
pathway_list <- unique(top_pathways$`GO/Pathway/Gene Set/Gene Name`)
filtered_data <- data_modules %>% filter(`GO/Pathway/Gene Set/Gene Name` %in% pathway_list)

# Add a new column to represent the range for the error bars
filtered_data <- filtered_data %>%
  mutate(lower = `BMD Median` - `BMD Standard Deviation`,
         upper = `BMD Median` + `BMD Standard Deviation`)

# Plot the BMD Median trends
p_3 <- ggplot(filtered_data, aes(x = timepoint, y = `BMD Median`, color = `GO/Pathway/Gene Set/Gene Name`)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  labs(title = "BMD Median Trends of Modules Over Time",
       x = "Timepoint",
       y = "BMD Median",
       color = "Modules") +
  theme_minimal() +
  facet_wrap(~`GO/Pathway/Gene Set/Gene Name`) +
  theme(legend.position = "none",
        strip.text = element_text(size = 8)) # Optional: wrap text

# Save the plot with desired dimensions
ggsave("BMD_Median_Trends_Modules.png", plot = p_3, width = 12, height = 8)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Extract timepoints and BMD SD for Modules_data
modules_data <- Modules_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.numeric(gsub(".*_(\\d+)countdata.*", "\\1", Analysis))) %>%
  select(timepoint, BMD_SD = `BMD Standard Deviation`) %>%
  mutate(Pathway = "Modules")

# Extract timepoints and BMD SD for Wiki_data
wiki_data <- Wiki_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.numeric(gsub(".*_(\\d+)countdata.*", "\\1", Analysis))) %>%
  select(timepoint, BMD_SD = `BMD Standard Deviation`) %>%
  mutate(Pathway = "Wiki")

# Extract timepoints and BMD SD for KEGG_data
kegg_data <- KEGG_data %>%
  separate(Analysis, into = c("compound", "timepoint"), sep = "_\\d+countdata", remove = FALSE) %>%
  mutate(timepoint = as.numeric(gsub(".*_(\\d+)countdata.*", "\\1", Analysis))) %>%
  select(timepoint, BMD_SD = `BMD Standard Deviation`) %>%
  mutate(Pathway = "KEGG")

# Combine the data
combined_sd <- bind_rows(modules_data, wiki_data, kegg_data)

# Ensure the interaction of Pathway and timepoint is used for the boxplot
combined_sd$Group <- interaction(combined_sd$Pathway, combined_sd$timepoint)

# Generate the group labels for the boxplot
group_labels <- paste(combined_sd$Pathway, combined_sd$timepoint, sep = "_")

# Define colors for different timepoints
timepoint_colors <- c("16" = "red", "24" = "blue", "48" = "green", "72" = "purple")

# Extract timepoints from group labels
timepoints <- sapply(strsplit(unique(group_labels), "_"), function(x) x[2])

# Assign colors based on timepoints
boxplot_colors <- timepoint_colors[timepoints]

# Plot the boxplot with further adjusted margins
par(mar = c(12, 5, 4, 2) + 0.1)  # Further increase the bottom margin
par(mgp = c(3, 2, 0))  # Adjust the position of axis labels

boxplot(BMD_SD ~ Group, 
        data = combined_sd,
        main = "Comparison of BMD Standard Deviation Across Timepoints",
        xlab = "",  # Leave xlab empty
        ylab = "BMD Standard Deviation",
        las = 2,  # Rotate x-axis labels for better readability
        names = unique(group_labels),
        col = boxplot_colors)  # Color the boxes according to timepoints

# Add a custom x-axis label using mtext
mtext("Pathway and Timepoint", side = 1, line = 6)  # Adjust 'line' to move the label down


```

```{r}
Gene16_data <- read_delim(file.path(input_dir, "Gene16_filtered.txt"), skip = 43, delim = "\t")
Gene24_data <- read_delim(file.path(input_dir, "Gene24_filtered.txt"), skip = 39, delim = "\t")
Gene48_data <- read_delim(file.path(input_dir, "Gene48_filtered.txt"), skip = 38, delim = "\t")
Gene72_data <- read_delim(file.path(input_dir, "Gene72_filtered.txt"), skip = 39, delim = "\t")

# Extract the 'Best BMD Median' column
# Replace 'Best BMD Median' with the actual name of your column if different
BMD_Median_72 <- Gene72_data$`Best BMD`

# Calculate the 5th percentile
percentile_5th_72 <- quantile(BMD_Median_72, 0.05, na.rm = TRUE)

# Print the result
print(percentile_5th_72)

BMD_Median_48 <- Gene48_data$`Best BMD`

# Calculate the 5th percentile
percentile_5th_48 <- quantile(BMD_Median_48, 0.05, na.rm = TRUE)

# Print the result
print(percentile_5th_48)

BMD_Median_24 <- Gene24_data$`Best BMD`

# Calculate the 5th percentile
percentile_5th_24 <- quantile(BMD_Median_24, 0.05, na.rm = TRUE)

# Print the result
print(percentile_5th_24)

BMD_Median_16 <- Gene16_data$`Best BMD`

# Calculate the 5th percentile
percentile_5th_16 <- quantile(BMD_Median_16, 0.05, na.rm = TRUE)

# Print the result
print(percentile_5th_16)

```

